---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# TabICL2 (WiP)

<!-- badges: start -->

[![R build status](https://github.com/cregouby/tabicl2/workflows/R-CMD-check/badge.svg)](https://github.com/cregouby/tabicl2/actions) [![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html) [![CRAN status](https://www.r-pkg.org/badges/version/tabicl2)](https://CRAN.R-project.org/package=tabicl2) [![](https://cranlogs.r-pkg.org/badges/tabicl2)](https://cran.r-project.org/package=tabicl2) [![Discord](https://img.shields.io/discord/837019024499277855?logo=discord)](https://discord.com/invite/s3D5cKhBkx)

<!-- badges: end -->

**TabICL2** is an **Work in Progress** R implementation of [TabICLv2: A better, faster, scalable, and open tabular foundation model](https://arxiv.org/abs/2602.11139) [(Jingang Qu, David Holzmüller, Gaël Varoquaux, Marine Le Morvan)](https://doi.org/10.48550/arXiv.2602.11139).\

**TabICL2** is a transformer-based foundation model for tabular data using **in-context learning**. Unlike traditional supervised learning that requires model fine-tuning, TabICL learns patterns from labeled examples in-context—similar to how GPT processes text, but for tabular data—to make predictions without additional training.

## Key Features

- **In-Context Learning**: Make predictions on new data by learning from labeled examples in the prompt, without fine-tuning
- **Universal Architecture**: Single model handles both classification and regression tasks
- **Scalable Classification**: Hierarchical classification automatically handles datasets with thousands of classes
- **Efficient Inference**: KV-caching and representation caching for fast batch predictions
- **Memory Management**: Automatic offloading to CPU or disk for large-scale inference
- **Distribution-Aware Embeddings**: Column embeddings capture statistical properties of features using set transformers
- **Flexible Attention**: Scalable softmax (SSMax) variants prevent attention saturation on long sequences

## Installation

You can install the development version of TabICL2 from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("cregouby/TabICL2")
```

## Quick Start

### Classification Example

```{r classification_example, eval=FALSE}
library(tabicl2)
library(torch)

# Create a TabICL model for classification
model <- TabICL(
  max_classes = 10,           # Maximum classes (0 for regression)
  num_quantiles = 10,         # Output dimension
  col_embed_dim = 64,         # Column embedding dimension
  row_d_model = 128,          # Row representation dimension
  icl_d_model = 256,          # ICL transformer dimension
  col_num_blocks = 2,         # Column transformer blocks
  row_num_blocks = 2,         # Row transformer blocks
  icl_num_blocks = 4,         # ICL transformer blocks
  nhead = 8                   # Attention heads
)

# Prepare your data
# X: (batch, total_rows, features) - training + test rows concatenated
# y_train: (batch, train_rows) - labels for training portion (0-based indices)
batch_size <- 4
train_size <- 20
test_size <- 10
n_features <- 15

X <- torch_randn(batch_size, train_size + test_size, n_features)
y_train <- torch_randint(0L, 5L, c(batch_size, train_size))  # 5 classes

# Make predictions on test rows using training examples as context
model$eval()
with_no_grad({
  predictions <- model(X, y_train, return_logits = FALSE)
  # predictions: (batch, test_size, num_classes) with probabilities
})

# Get predicted classes
predicted_classes <- predictions$argmax(dim = -1)
print(predicted_classes)
```

### Regression Example

```{r regression_example, eval=FALSE}
library(tabicl2)
library(torch)

# For regression, set max_classes = 0
model <- TabICL(
  max_classes = 0,            # 0 = regression mode
  num_quantiles = 9,          # Predict 9 quantiles
  col_embed_dim = 64,
  row_d_model = 128,
  icl_d_model = 256,
  col_num_blocks = 2,
  row_num_blocks = 2,
  icl_num_blocks = 4,
  nhead = 8
)

# Prepare data
X <- torch_randn(batch_size, train_size + test_size, n_features)
y_train <- torch_randn(batch_size, train_size)  # Continuous targets

# Predict quantiles for test rows
model$eval()
with_no_grad({
  quantile_preds <- model(X, y_train)
  # quantile_preds: (batch, test_size, num_quantiles)
})

# Convert quantiles to mean and variance
predictions <- QuantileToDistribution(quantile_preds, y_train)
# predictions$mean: (batch, test_size, 1)
# predictions$variance: (batch, test_size, 1)
```

## Advanced Features

### KV Caching for Fast Inference

When making multiple predictions on the same training context, use caching to avoid redundant computation:

```{r caching_example, eval=FALSE}
# Store cache on first forward pass
model$forward_with_cache(
  X = X_train,              # Training data only
  y_train = y_train,
  store_cache = TRUE,
  cache_mode = "kv"         # Cache key-value projections
)

# Reuse cache for subsequent test batches
for (test_batch in test_batches) {
  preds <- model$forward_with_cache(
    X = test_batch,
    use_cache = TRUE,
    num_classes = 5
  )
}

# Clear cache when done
model$clear_cache()
```

**Cache Modes:**
- `"kv"`: Cache key-value projections (fastest, more memory)
- `"repr"`: Cache row representations only (~24x less memory for ICL part)

### Memory-Efficient Inference

For large-scale inference, configure automatic offloading:

```{r inference_config, eval=FALSE}
# Configure inference settings
inf_config <- InferenceConfig(
  auto_batch = TRUE,          # Automatic batching
  batch_size = 32,
  offload = OFFLOAD_AUTO,     # Automatic memory management
  min_memory_gb = 4.0         # Keep 4GB GPU memory available
)

# Use configuration during inference
predictions <- model$inference_forward(
  X, y_train,
  inference_config = inf_config
)
```

**Offload Options:**
- `OFFLOAD_GPU`: Keep all on GPU (fastest)
- `OFFLOAD_CPU`: Offload to CPU pinned memory
- `OFFLOAD_DISK`: Offload to memory-mapped files
- `OFFLOAD_AUTO`: Automatically choose based on available memory

### Hierarchical Classification

TabICL automatically handles datasets with more classes than `max_classes` using hierarchical classification:

```{r hierarchical_example, eval=FALSE}
# Model with max_classes = 10
model <- TabICL(max_classes = 10, num_quantiles = 10, ...)

# Dataset with 100 classes - automatically uses hierarchy
y_train_large <- torch_randint(0L, 100L, c(batch_size, train_size))

model$eval()
predictions <- model(X, y_train_large, return_logits = FALSE)
# predictions: (batch, test_size, 100) probabilities via hierarchical ensembling
```

## Architecture

TabICL processes tabular data through three sequential stages:

### 1. Column-wise Embedding (`ColEmbedding`)

Creates distribution-aware embeddings for each column using set transformers with induced self-attention:

- Maps scalar cells to high-dimensional embeddings
- Captures statistical regularities within columns
- Uses shared parameters across all features
- Supports feature grouping and target-aware embeddings
- Optional affine transformations for final embeddings

### 2. Row-wise Interaction (`RowInteraction`)

Captures interactions between features within each row using standard transformer blocks:

- Applies rotary position embeddings (RoPE) to encode feature positions
- Uses learnable CLS tokens to aggregate feature information
- Outputs: concatenated CLS token embeddings as row representations

### 3. Dataset-wise In-Context Learning (`ICLearning`)

Learns from labeled training examples in-context to predict on test examples:

- Input sequence: `[train_row_1, ..., train_row_n, test_row_1, ..., test_row_m]`
- Training rows are augmented with label embeddings
- Uses transformer architecture with scalable softmax
- For classification: outputs logits/probabilities for each class
- For regression: outputs quantile predictions

## How It Works

```r
# Training mode: Learn to predict from in-context examples
model$train()
loss <- compute_loss(model(X_batch, y_train_batch), y_true)

# Inference mode: Use learned patterns to make predictions
model$eval()
predictions <- model(X_new, y_context)
```

**Key Insight**: The model is trained on many different tabular datasets. At inference time, it uses the labeled examples you provide (the "context") to understand the current task and make predictions on unlabeled test rows—no fine-tuning required.

## Training vs Inference

- **Training mode** (`model$train()`): Learns to perform in-context learning across diverse tabular datasets
- **Evaluation mode** (`model$eval()`): Uses learned in-context learning ability to make predictions

The model must be pre-trained on a large corpus of tabular datasets before it can perform effective in-context learning at inference time.

## Performance Tips

1. **Use caching** when making multiple predictions with the same training context
2. **Enable auto_batch** for variable-length sequences in the same batch
3. **Configure offloading** when GPU memory is limited
4. **Use gradient checkpointing** (`recompute = TRUE`) during training to save memory
5. **Adjust temperature** (`softmax_temperature`) to calibrate prediction confidence

## References

- **TabICL**: [A Tabular Foundation Model for In-Context Learning on Large Data](https://arxiv.org/abs/2502.05564)
- **TabPFNv2**: [Accurate predictions on small data with a tabular foundation model](https://github.com/automl/TabPFN)
- **Set Transformers**: [Lee et al., 2019](https://arxiv.org/abs/1810.00825)
- **In-Context Learning**: Learning from examples in the prompt without parameter updates

## Citation

If you use TabICL2 in your research, please cite:

```bibtex
@software{tabicl2,
  title = {TabICL2: Transformer-based In-Context Learning for Tabular Data},
  author = {Your Name},
  year = {2026},
  url = {https://github.com/cregouby/TabICL2}
}
```

## License

See LICENSE file for details.

